The Design and Implementation of the Synchronous Language \CEU

### 2016-04-01 paper.tex
...

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution which is amenable to a simple temporal analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class synchronized timers with a dedicated syntax.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing the practical aspect of our clean semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

\CEU aims to offer a simpler semantics than Esterel without giving up its 
tractable synchronous concurrency model:

\CEU is a synchronous language inspired by Esterel which aims to offer a 
simpler semantics and more fine-grained control over the program execution.
%
\CEU employs an event-triggered notion of time that allows for a simple
reasoning about reactions in parallel and enables compile-time analysis to 
produce deterministic and concurrency-safe programs.
%
We discuss the particularities of our design in comparison to Esterel, such as 
stack-based internal events, temporal analysis for concurrent statements, safe 
integration with \C, and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.

Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution which is amenable to a simple temporal analysis that
improves safety.

Contributions:
%
\begin{itemize}
\item \textbf{Event-triggered notion of time bound to the semantics of the 
language}, which allows ....
\item \textbf{Deterministic and fine-grained intra-reaction execution and 
communication}, lexical, stack-based internal events
\item \textbf{Static temporal analysis for concurrent statements}, which
      prevent logical race conditions on shared variables and system calls.
\item \textbf{Safe integration with \C}, with enforced resource clean up.
\item \textbf{First-class timers} with dedicated syntax and automatic 
synchronization.
\end{itemize}

Limitations:
%
\begin{itemize}
\item purely reactive execution,
\item bounded-only reactions,
\item sequential
\end{itemize}
%
We believe that a simple-tractable synchronous core
and extend it with memory-isolated parallelizabe asynchronous execution.

    - permeates all aspects
    - As far as we know, \CEU is the first Esterel-based language to HAVE?
      queue-based/unique in the semantics of the language and not in the
      library
    - stack
        - deterministic intra-reaction execution
    - imposed intra-reaction execution
        - for inter-reaction commm
        - for intra-reaction commm
        - for stmts

@@

The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
through a queue and are dispatched sequentially and atomically to prevent race 
conditions.
%
We believe that for software design, this approach is more familiar to 
programmers and simplifies the reasoning about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis that enables safe shared-memory concurrency.

>>>>

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution which is amenable to a simple temporal analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing the practical aspect of our clean semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

\CEU aims to offer a deterministic and simpler semantics than Esterel with a 
more fine-grained control over the program execution.
%
The list that follows enumerate the contributions of our design in this 
direction:
%
\begin{itemize}
%
\item Event-triggered notion of time bound to the semantics of the language.
Event-driven programming is popular in many domains, such as server and GUI 
development.
We believe that programmers are more familiar with this notion of time, which 
simplifies the reasoning about concurrency.
For instance, the uniqueness of external events is a prerequisite for the 
temporal analysis of \CEU.
%
\item Deterministic intra-reaction execution and communication.
Determinism in \CEU is all-inclusive and does not depend on additional levels 
of static analysis.
It encompasses the whole language, including memory accesses, system calls, and 
the expressive stack-based internal events.
For instance, programmers can always figure out which statement executes next, 
making runtime analysis and debugging easier.
%
\item Static temporal analysis for concurrent statements.
Although execution is deterministic, the compiler of \CEU still advises about 
suspicious statements reacting concurrently to the same event.
This way, the complier forces the programmer to serialize the statements 
accordingly.
%
\item Safe integration with \C.
\CEU forces the programmer to deal with pointers representing resources through 
finalization clauses.
When dealing with concurrent system calls, programmers can provide the compiler 
with annotations to remove false positives from the temporal analysis or to 
allow non commutative concurrent behavior.
%
\item First-class timers with dedicated syntax and automatic synchronization.
Given the omnipresence of timers in embedded systems, a dedicated syntax can 
simplify the development and readability of programs.
Furthermore, automatic synchronization releases the programmer from the burden 
of accumulating timers in sequence and comparing timers in parallel.
%
\end{itemize}

The design decisions XXX, raise some important limitations as follows:
%
\begin{itemize}
\item Purely reactive execution.
    hardware
A limitation of event-triggered execution is that all program behavior must be purely
reactive, given that no code can execute in the absence of inputs. Tick-triggered execu-
tion allows for active behavior, since code can execute regulary on every tick. Although
it supports asynchronous active execution [Sant’Anna et al. 2012], we still consider
 ́ as purely reactive since its synchronous core cannot express active behavior.
\item bounded-only reactions,
\item sequential
\end{itemize}
%
We believe that a simple-tractable synchronous core
and extend it with memory-isolated parallelizabe asynchronous execution.

    - permeates all aspects
    - As far as we know, \CEU is the first Esterel-based language to HAVE?
      queue-based/unique in the semantics of the language and not in the
      library

@@

The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
through a queue and are dispatched sequentially and atomically to prevent race 
conditions.
%
%We believe that for software design, this approach is more familiar to 
%programmers and simplifies the reasoning about concurrency.
%For instance, the uniqueness of external events in \CEU is a prerequisite for 
%the temporal analysis that enables safe shared-memory concurrency.
